Toward \earlywarning\ detection of gravitational waves from compact binary coalescence

### 2011-10-12 article.tex
Addressed referee's comment 4
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1392 ccdd05d5-56c1-461a-bf98-85f6876523b4

\begin{figure}[h]
\plotone{f1}
\caption{\label{fig:earlywarning}Expected number of \NS--\NS\
sources that could be detectable by Advanced \LIGO\ a given number of seconds
before coalescence.  The heavy solid line corresponds to the most probable yearly rate
estimate from~\citet{Abadie:2010p10836}.  The shaded region represents the 5\% to 95\% confidence interval
arising from substantial uncertainty in predicted event rates.}
\end{figure}
%
Before the \GW\ signal leaves the detection band, we can imagine examining the
signal to noise ratio (\SNR) accumulated up to that point and if it is
already significant, release an alert immediately, trading \SNR\ and sky
localization accuracy for pre-merger detection.
Figure~\ref{fig:earlywarning} shows projected early detectability rates for
\NS--\NS\ binaries in Advanced \LIGO\ assuming the anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise} and
\NS--\NS\ merger rates estimated in \citet{Abadie:2010p10836}.  The merger rates
have substantial measurement uncertainty due to the small sample of known double pulsar systems that will merge within a Hubble time; they also have systematic uncertainty due sensitive dependence on the pulsar luminosity distribution function~\citep{KalogeraRates}.
The most probable estimates indicate that at a single-detector \SNR\ threshold of 8, we will
observe a total of 40~events~yr$^{-1}$; $\sim$10~yr$^{-1}$ will be detectable
within 10~s of merger and $\sim$5~yr$^{-1}$ will be detectable within 25~s of
merger if analysis can proceed with near zero latency.

@@

For concreteness and simplicity, consider an inspiral waveform in the
quadrapole approximation, for which the time-frequency relation~\citep{kidder1992, findchirppaper} is
%
\begin{equation} \label{eq:fgw}
%
f(t) = \frac{1}{\pi \mathcal{M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5$~$\upmu$s) and $t$ is the time relative to the
coalescence of the binary.
This monotonic time-frequency relationship allows us
to choose time slice boundaries that require substantially less bandwidth at
early times in the inspiral.

>>>>

\begin{figure}[h]
\plotone{f1}
\caption{\label{fig:earlywarning}Expected number of \NS--\NS\
sources that could be detectable by Advanced \LIGO\ a given number of seconds
before coalescence.  The heavy solid line corresponds to the most probable yearly rate
estimate from~\citet{Abadie:2010p10836}.  The shaded region represents the 5\% to 95\% confidence interval
arising from substantial uncertainty in predicted event rates.}
\end{figure}
%
Before the \GW\ signal leaves the detection band, we can imagine examining the
signal to noise ratio (\SNR) accumulated up to that point and if it is
already significant, release an alert immediately, trading \SNR\ and sky
localization accuracy for pre-merger detection.

In the quadrupole approximation, the instantaneous frequency of the GW inspiral signal is related to the time $t$ relative to coalescence ~\citep{kidder1992, findchirppaper} through
%
\begin{equation} \label{eq:fgw}
	f(t) = \frac{1}{\pi \mathcal{M}}
		\left[ \frac{5}{256}\frac{\mathcal{M}}{t} \right]^{3/8}
\end{equation}
%
where $\mathcal{M} = M^{2/5} \mu^{3/5}$ is the chirp mass of the binary in units of time ($G
M_\odot / c^3 \approx 5$~$\upmu$s), $M$ is the total mass, and $\mu$ is the reduced mass. 
The expected value of the single detector \SNR\ for an optimally oriented (source at detector's zenith or nadir, orbital plane face-on) inspiral source is~\citep{Abadie:2010p10836}
%
\begin{equation}
	\label{eq:expected-snr}
	\rho =
		\frac{\mathcal{M}^{5/6} c}{\pi^{2/3} D}
		\sqrt{
			\frac{5}{6} \int_{f_\mathrm{low}}^{f_\mathrm{high}}
			\frac{f^{-7/3}}{S(f)} \mathrm{d}f}
\end{equation}
%
where $D$ is the luminosity distance and $S(f)$ is the power spectral density of the detector noise.  $f_\mathrm{low}$ and $f_\mathrm{high}$ are low and high frequency limits of integration which may be chosen to extend across the entire bandwidth of the detector.  If we want to trigger at a time $t$ before merger, then we must cut off the SNR integration at $f_\mathrm{high} = f(t)$ with $f(t)$ given by equation~(\ref{eq:fgw}) above.

Figure~\ref{fig:earlywarning} shows projected early detectability rates for
\NS--\NS\ binaries in Advanced \LIGO\ assuming the anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise} and
\NS--\NS\ merger rates estimated in \citet{Abadie:2010p10836}.  The merger rates
have substantial measurement uncertainty due to the small sample of known double pulsar systems that will merge within a Hubble time; they also have systematic uncertainty due sensitive dependence on the pulsar luminosity distribution function~\citep{KalogeraRates}.
The most probable estimates indicate that at a single-detector \SNR\ threshold of 8, we will
observe a total of 40~events~yr$^{-1}$; $\sim$10~yr$^{-1}$ will be detectable
within 10~s of merger and $\sim$5~yr$^{-1}$ will be detectable within 25~s of
merger if analysis can proceed with near zero latency.

@@

For concreteness and simplicity, consider an inspiral waveform in the
quadrapole approximation, for which the time-frequency relation is given by equation~(\ref{eq:fgw}).
This monotonic time-frequency relationship allows us
to choose time slice boundaries that require substantially less bandwidth at
early times in the inspiral.
